---
title: "Barclays Premier League Simulation"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}
library(tidyverse)
```

#Read in Premier Leage Dataframe
```{r}
soccer<-read_csv("/Users/jake/Downloads/spi_global_rankings.csv")
```

#Look at some summary statistics
```{r}
mean.off<-mean(soccer$off)
mean.def<-mean(soccer$def)
lmean.off<-mean(log(soccer$off))
lmean.def<-mean(log(soccer$def) ) 
mean.off
mean.def
```

#Filter dataframe to only include teams from Barclay's Premier League, the League we will be simulating
```{r}
premLeague<-subset(soccer,league=="Barclays Premier League")
premTeams<-premLeague$name
rownames(premLeague)<-premTeams
df.prem<-premLeague[,c("off","def")]
rownames(df.prem)<-premTeams
df.prem$premTeams<-premTeams
df.prem<-with(df.prem,df.prem[order(premTeams),])
df.prem
```

#Look at new summary statistics.
```{r}
lmean.def<- log(mean(df.prem$def))
lmean.off<- log(mean(df.prem$off))               
df.prem["alpha"]<-log(df.prem["off"])-lmean.def
df.prem["delta"]<-lmean.off-log(df.prem["def"])
#View(df.prem)
```

#Build alpha and delta lists for poisson distribution
```{r}
alphaList<-df.prem$alpha
deltaList<-df.prem$delta
names(alphaList)<-df.prem$premTeams
names(deltaList)<-df.prem$premTeams
alphaList["Everton"]
```

#Create function that will return a score for a game based on the poisson distribution
```{r}
draw.score<-function(team1,team2){
    c(
        rpois(1,exp(alphaList[team1]-deltaList[team2])),
  rpois(1,exp(alphaList[team2]-deltaList[team1]))
    )
}
draw.score("Liverpool","Arsenal")
```

#Get all of the games played in a season
```{r}
#install.packages('gtools')
library('gtools')
# All possible matches in a season
allMatches<-permutations(20, 2, v=premTeams,repeats.allowed=FALSE)
colnames(allMatches)<-c("home","away")
head(allMatches,9)
length(allMatches)
```


#Write a function that uses all of the games in a season and scores generator to simulate one season of games
```{r}
generate_scores_matrix<-function(){
  ScoresMatrix <- as.data.frame(matrix(nrow=nrow(allMatches),  ncol=4))
  for (ii in 1:nrow(allMatches)  ) {
       ScoresMatrix[ii,1:2]=allMatches[ii,]
       ScoresMatrix[ii,3:4]= draw.score(allMatches[ii,"home"],allMatches[ii,"away"] )  
  }
  colnames(ScoresMatrix)<-c("home.team","away.team","home.score","away.score")
  ScoresMatrix$home.points<-0
  ScoresMatrix$home.points[ScoresMatrix$home.score>ScoresMatrix$away.score]<-3
  ScoresMatrix$home.points[ScoresMatrix$home.score==ScoresMatrix$away.score]<-1
  
  ScoresMatrix$away.points<-0
  ScoresMatrix$away.points[ScoresMatrix$away.score>ScoresMatrix$home.score]<-3
  ScoresMatrix$away.points[ScoresMatrix$away.score==ScoresMatrix$home.score]<-1
  ScoresMatrix$score.dif<-ScoresMatrix$home.score-ScoresMatrix$away.score
  
  sum_scoresdif<-rep(999,20)
  goals_scored<-rep(999,20)
  l<-1
  
  for(name in premTeams){
  tempteam<-ScoresMatrix%>%filter(home.team==name)
  sum_scoresdif[l]<-sum(tempteam$score.dif)
  goals_scored[l]<-sum(tempteam$home.score)
  tempteam<-ScoresMatrix%>%filter(away.team==name)
  sum_scoresdif[l]<-sum_scoresdif[l] - sum(tempteam$score.dif)
  goals_scored[l]<-goals_scored[l] + sum(tempteam$away.score)
  l<-l+1
}
  return(list(ScoresMatrix, sum_scoresdif, goals_scored))

}

generate_scores_matrix()[[1]]
```


#Assign points to teams and corresponding monetary gains based on final rank 
```{r}
money<-c(220,210,200,180,160,150,140,125,120,117,116,115,113,110,108,105,100,51,50,49)


generate_win_list<-function(){
  # Example scores through the entire season
  scoreslist<-generate_scores_matrix()
  ScoresMatrix<-scoreslist[[1]]
  places<-as.data.frame(matrix(nrow=20, ncol=2))
  z<-1
  for(j in premTeams){
    tempteam<-ScoresMatrix%>%filter(home.team==j | away.team==j)
    places[z,1]<-j
    wincount<-0
    for(k in 1:nrow(tempteam)){
      if(tempteam$home.team[k]==j){
        wincount<-wincount+tempteam$home.points[k]     
      }
      else{
        wincount<-wincount+tempteam$away.points[k]
      }
    }
    places[z,2]<-wincount
    z<-z+1
  }
  places$sum_goals<-scoreslist[[2]]
  places$goals_scored<-scoreslist[[3]]
  places<-places[order(-places$V2, -places$sum_goals, -places$goals_scored), ]
  places$place<-1:20
  places$top7<-0
  places$top7[places$place<8]<-1
  places$middle10<-0
  places$middle10[places$place>=8 & places$place<=17]<-1
  places$bottom3<-0
  places$bottom3[places$place>17]<-1
  places$money<-money
  
  return(places)
}
generate_win_list()
```

#Simulate 100 seasons
```{r}
ordered_list<-generate_win_list()
for(i in 1:100){
  temp_ordered_list<-generate_win_list()
  ordered_list<-rbind(ordered_list,temp_ordered_list)
  print(i)
}
```

#What were the average earnings for teams over 100 seasons
```{r}
ordered_list$money<-rep(money, 101)
avg_place<-ordered_list%>%group_by(V1)%>%summarise(mean_place=mean(place), med_place=median(place),mean_top7=mean(top7), mean_middle10=mean(middle10), mean_bottom3=mean(bottom3), sd_place=sd(place), mean_money=mean(money), sd_money=sd(money))
avg_place<-avg_place[order(-avg_place$sd_money),]
avg_place$final_place<-1:20
ordered_list
avg_place
```


```{r}
ggplot(avg_place, aes(x=reorder(V1,-mean_place), y=mean_place)) + geom_col() +coord_flip() +theme_bw() +labs(x="Team Order", y="Expected Ranking", title="Ranked Team", subtitle="Barclay's Premier League")
```


```{r}
ggplot(ordered_list, aes(x=reorder(V1,money), y=money))+geom_boxplot() +coord_flip() +theme_bw() +labs(x="Team", y="Financial Rewards(in Millions)", title="Variability of Earnings for Every Team", subtitle="Barclay's Premier League")
```   

#How much money does 1 additional win grant each team
```{r}
added_money<-as.data.frame(matrix(nrow=20, ncol=5))
colnames(added_money)<-c("Team.Name", "mean_pre3", "mean_post3", "sd_pre3", "sd_post3")

x<-1

for(name in premTeams){
  added_money[x,1]<-name
  ordered_list1<-ordered_list
  
  ordered_list1$V2<-as.numeric(ordered_list1$V2)
  
  ordered_list1$V2[ordered_list1$V1==name]<-ordered_list1$V2[ordered_list1$V1==name]+3
  
  ordered_list1
  
  num<-21
  
  reordered_list<-ordered_list1[1:20,]
  reordered_list<-reordered_list[order(-reordered_list$V2, -reordered_list$sum_goals, -reordered_list$goals_scored), ]
  reordered_list$place<-1:20
  
  
  for(i in 2:101){
    num2<-num+19
    league<-ordered_list1[num:num2,]
    league<-league[order(-league$V2, -league$sum_goals, -league$goals_scored), ]
    league$place<-1:20
    reordered_list<-rbind(reordered_list, league)
    num<-i*20+1
  }
  
  reordered_list$money<-rep(money, 101)
  
  added_money[x,2]<-mean(ordered_list$money[ordered_list$V1==name])
  added_money[x,3]<-mean(reordered_list$money[reordered_list$V1==name])
  added_money[x,4]<-sd(ordered_list$money[ordered_list$V1==name])
  added_money[x,5]<-sd(reordered_list$money[reordered_list$V1==name])
  x<-x+1
}

added_money$money_added<-added_money$mean_post3-added_money$mean_pre3
added_money
```
```{r}
ggplot(added_money, aes(x=reorder(Team.Name,money_added), y=money_added))+geom_col(fill="orange")+coord_flip()
```


#Calculate 10% increase in defense for each team
```{r}
f<- function(x, df, sum){
  0.9*sum - (exp(df$alpha[1]-x)+exp(df$alpha[2]-x)+exp(df$alpha[3]-x)+exp(df$alpha[4]-x)+exp(df$alpha[5]-x)+exp(df$alpha[6]-x)+exp(df$alpha[7]-x)+exp(df$alpha[8]-x)+exp(df$alpha[9]-x)+exp(df$alpha[10]-x)+exp(df$alpha[11]-x)+exp(df$alpha[12]-x)+exp(df$alpha[13]-x)+exp(df$alpha[14]-x)+exp(df$alpha[15]-x)+exp(df$alpha[16]-x)+exp(df$alpha[17]-x)+exp(df$alpha[18]-x)+exp(df$alpha[19]-x))
}


#delta
deltas<-df.prem$delta
new_deltas<-rep(999,20)
x<-1
for(name in premTeams){
  tempdf<-cbind(df.prem$premTeams, data_frame(c(rep(deltas[x], 20))), df.prem$alpha)
  colnames(tempdf)=c("team", "team_delta","alpha")
  tempdf<-tempdf[-x, ]
  tempdf$exp_diff<- exp(tempdf$alpha-tempdf$team_delta)
  sum<- sum(tempdf$exp_diff)
  new_deltas[x]<-newton.rhapson(f, tempdf, sum, 0,output=TRUE)
  x<-x+1
}

new_deltas
```

#simulate 100 seasons for each team 10% increase in defense
```{r}
x<-0
defense_money<-rep(0,20)
for(team in premTeams){
  x<-x+1
  alphaList<-df.prem$alpha
  names(alphaList)<-df.prem$premTeams
  deltaList<-df.prem$delta
  names(deltaList)<-df.prem$premTeams
  deltaList[[x]]<-new_deltas[x]
  ordered_list_def<-generate_win_list()
  for(i in 1:100){
    temp_ordered_list<-generate_win_list()
  ordered_list_def<-rbind(ordered_list_def,temp_ordered_list)
  }
  avg_place_def<-ordered_list_def%>%group_by(V1)%>%summarise(mean_place=mean(place), med_place=median(place),mean_top7=mean(top7), mean_middle10=mean(middle10), mean_bottom3=mean(bottom3), sd_place=sd(place), mean_money=mean(money), sd_money=sd(money))
  #avg_place_def<-avg_place_def[order(-avg_place_def$mean_money),]
  #print(ordered_list)
  #print(avg_place)
  print(team)
  defense_money[x]<-avg_place_def[x,8]
}

defense_money
```

#Calculate 10% increase in offense for each team
```{r}
f<- function(x, df, sum){
  1.1*sum - (exp(x-df$delta[1])+exp(x-df$delta[2])+exp(x-df$delta[3])+exp(x-df$delta[4])+exp(x-df$delta[5])+exp(x-df$delta[6])+exp(x-df$delta[7])+exp(x-df$delta[8])+exp(x-df$delta[9])+exp(x-df$delta[10])+exp(x-df$delta[11])+exp(x-df$delta[12])+exp(x-df$delta[13])+exp(x-df$delta[14])+exp(x-df$delta[15])+exp(x-df$delta[16])+exp(x-df$delta[17])+exp(x-df$delta[18])+exp(x-df$delta[19]))
}

#man city
alphas<-df.prem$alpha
new_alphas<-rep(999,20)
x<-1
for(name in premTeams){
  tempdf<-cbind(df.prem$premTeams, data_frame(c(rep(alphas[x], 20))), df.prem$delta)
  colnames(tempdf)=c("team", "team_alpha","delta")
  tempdf<-tempdf[-x, ]
  tempdf$exp_diff<- exp(tempdf$team_alpha-tempdf$delta)
  sum<- sum(tempdf$exp_diff)
  new_alphas[x]<-newton.rhapson(f, tempdf, sum, 0,output=TRUE)
  x<-x+1
}

new_alphas
```

#simulate 100 seasons for each team having 10% increased offense
```{r}
x<-0
offense_money<-rep(0,20)
for(team in premTeams){
  x<-x+1
  alphaList<-df.prem$alpha
  names(alphaList)<-df.prem$premTeams
  deltaList<-df.prem$delta
  names(deltaList)<-df.prem$premTeams
  alphaList[[x]]<-new_alphas[x]
  ordered_list_off<-generate_win_list()
  for(i in 1:100){
    temp_ordered_list<-generate_win_list()
    ordered_list_off<-rbind(ordered_list_off,temp_ordered_list)
  }
  avg_place_off<-ordered_list_off%>%group_by(V1)%>%summarise(mean_place=mean(place), med_place=median(place),mean_top7=mean(top7), mean_middle10=mean(middle10), mean_bottom3=mean(bottom3), sd_place=sd(place), mean_money=mean(money), sd_money=sd(money))
  #print(ordered_list)
  #print(avg_place)
  print(team)
  print(avg_place_off)
  offense_money[x]<-avg_place_off[x,8]
}

offense_money
```

```{r}
alphaList<-df.prem$alpha
names(alphaList)<-df.prem$premTeams
alphaList
alphaList[[1]]<-new_alphas[1]
alphaList
```

#Do teams gain (monetarily) more from 10% offense increase or 10% defense increase
```{r}
#names(offense_money)<-premTeams
#offense_money<-as.data.frame(offense_money)
offense_money1<-unlist(offense_money)
offense_money1
defense_money1<-unlist(defense_money)
avg_place<-with(avg_place, avg_place[order(V1),])

avg_place$offense_money<-offense_money1
avg_place$money_invest_off<-avg_place$offense_money-avg_place$mean_money
avg_place$defense_money<-defense_money1
avg_place$money_invest_def<-avg_place$defense_money-avg_place$mean_money
avg_place
```

```{r}
ggplot(avg_place, aes(x=reorder(V1, money_invest_def)))+geom_col(aes(y=money_invest_def), fill="blue",position = 'dodge')+geom_col(aes(y=money_invest_off), fill="orange",position = 'dodge')+ theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```


